Relational Programming in Smalltalk

### 2018-09-08 ESUG2018/esug.tex
Add some slides.

@@

\begin{frame}[fragile]
\frametitle{Resolution by \textit{Unification}}
let $G$ be a set of equations, unification rules are
\begin{description}
\item[delete] $G \cup \lbrace t = t \rbrace \rightarrow G$
\item[decompose] $G \cup \lbrace f(s_{0}, \ldots, s_{k}) = f(t_{0}, \ldots, t_{k})\rbrace$ entails
$$G \cup \lbrace s_{0}=t_{0},\ldots, s_{k}=t_{k} \rbrace$$
\item[conflict] if $f\neq g \vee k\neq m$ then $$G \cup \lbrace f(s_{0}, \ldots, s_{k}) = g(t_{0}, \ldots, t_{m})\rbrace \rightarrow \,\perp$$
\item[eliminate] if $x \not\in vars(t)$ and $x \in vars(G)$ then $$G \cup \lbrace x = t\rbrace \rightarrow G\lbrace x \mapsto t\rbrace \cup \left\lbrace x \triangleq t\right\rbrace $$
\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = t(s_{0}, \ldots, s_{k}\rbrace \rightarrow \,\perp$$
\end{description}
%without *occur checks*, generating a substitution $\theta$ is a *recursive enumerable* problem
\end{frame}

@@

@@

\begin{frame}[fragile]
\frametitle{Unifier}
Let $\mathcal{D}$ be the set of \textit{Dyck paths} and let $\leadsto$ be the \textit{CFG}
$$
\left\lbrace\begin{array}{l}
\leadsto = \varepsilon \\
\leadsto = ( \leadsto ) \leadsto \\
\end{array}\right\rbrace
$$
question: enumerate $\mathcal{D}$ using $\leadsto$.
\begin{minted}[fontsize=\footnotesize]{smalltalk}
dycko: alpha
    ^ Goal cond e
        if: alpha nilo then: true asGoal;
        else: [ :beta :gamma | 
            (sexpTheory let: alpha 
                        be: ($( cons: beta) 
                        append: ($) cons: gamma)) & 
            ([ self dycko: beta ] eta & 
             [ self dycko: gamma ] eta) ] fresh
\end{minted}
\end{frame}

>>>>

\begin{frame}[fragile]
\frametitle{Main idea}
\begin{displaymath}
\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{entails}\quad P(1,2,3)
\end{displaymath}
can be expressed using either the \textit{imperative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
a := 1.
b := 2.
c := a + b.
Object assert: [ c = 3 ].
\end{minted}
or the \textit{functional style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
([ :a :b | a + b ] value: 1 value: 2) = 3
\end{minted}
or, finally, the \textit{declarative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
[ :a :b :c | a + b = c ] value: 1 value: 2 value: 3
\end{minted}

\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Resolution by \textit{Unification}}
let $G$ be a set of equations, unification rules are
\begin{description}
\item[delete] $G \cup \lbrace t = t \rbrace \rightarrow G$
\item[decompose] $G \cup \lbrace f(s_{0}, \ldots, s_{k}) = f(t_{0}, \ldots, t_{k})\rbrace$ entails
$$G \cup \lbrace s_{0}=t_{0},\ldots, s_{k}=t_{k} \rbrace$$
\item[conflict] if $f\neq g \vee k\neq m$ then $$G \cup \lbrace f(s_{0}, \ldots, s_{k}) = g(t_{0}, \ldots, t_{m})\rbrace \rightarrow \,\perp$$
\item[eliminate] if $x \not\in vars(t)$ and $x \in vars(G)$ then $$G \cup \lbrace x = t\rbrace \rightarrow G\lbrace x \mapsto t\rbrace \cup \left\lbrace x \triangleq t\right\rbrace $$
\item[occur check] if $x \in vars(f(s_{0},\ldots,s_{k}))$ then $$G \cup \lbrace x = t(s_{0}, \ldots, s_{k})\rbrace \rightarrow \,\perp$$
\end{description}
%without *occur checks*, generating a substitution $\theta$ is a *recursive enumerable* problem
\end{frame}

@@

A \textit{goal} is an object that responds to the \verb|#onState:| selector, it
receives a \textit{substitution} and returns a \verb|Chain| object of
substitutions.

@@

\begin{frame}[fragile]
\frametitle{Unifier}
Let $\mathcal{D}$ be the set of \textit{Dyck paths} and let $\leadsto$ be the \textit{CFG}
$$
\leadsto = \varepsilon \,\left|\, ( \leadsto ) \leadsto \right.
$$
where $\varepsilon$ is the empty string; question: enumerate $\mathcal{D}$ using $\leadsto$.
\begin{minted}[fontsize=\footnotesize]{smalltalk}
dycko: alpha
    ^ Goal cond e
        if: alpha nilo then: true asGoal;
        else: [ :beta :gamma | 
            (sexpTheory let: alpha 
                        be: ($( cons: beta) 
                        append: ($) cons: gamma)) & 
            ([ self dycko: beta ] eta & 
             [ self dycko: gamma ] eta) ] fresh
\end{minted}
\end{frame}
